Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-12-01 papers/basic-theory/paper/paper.tex
Patrick's edits. Many formatting changes. Improved intro + discuss

% Talon's custom commands
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*} Authors TBD,\authormark{2,*}, Rudolf Oldenbourg\authormark{2,*}, and Patrick La Rivi\`ere \authormark{1}}

\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, Illinois 60637, USA\\
\authormark{2}Publications Department, The Optical Society, 2010 Massachusetts Avenue NW, Washington, DC 20036, USA\\
\authormark{3}Currently with the Department of Electronic Journals, The Optical Society, 2010 Massachusetts Avenue NW, Washington, DC 20036, USA}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space
  operator, % with domain
  % $\mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$ and range $\mbb{L}_2(\mbb{R}^2)$
  and we express the operator in terms of four different sets of object-space
  basis functions. We show that the operator takes a particularly convenient
  form when expressed in a basis of spatial and spherical harmonics---a form we
  call the dipole spatio-angular transfer function. We demonstrate our formalism
  by analyzing a single-view fluorescence microscope without using the monopole
  or scalar approximations. We show that this imaging system has an angular band
  limit, and we demonstrate the value of the transfer function approach by
  efficiently simulating the imaging process with simple phantoms. Notably, we
  show that information about the out-of-plane orientation of ensembles of
  in-focus fluorophores is recorded by fluorescence microscopes. We discuss the
  implications of our analysis for all quantitative fluorescence microscopy
  studies and lay out a path towards a complete theory.
\end{abstract*}

\section{Introduction}

Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial distribution of fluorophores throughout a sample [cite].
While an unprocessed fluorescence micrograph reports the approximate
distribution of fluorophores throughout a sample, all microscopes are
diffraction limited, so the image is a blurred version of the true
fluorophore distribution.

Restoration techniques attempt to recover the true distribution of fluorophores
using the measured data and a model of the imaging process. A model of the
imaging process can be obtained theoretically (by mathematically modeling the
instrument under idealized conditions), experimentally (by measuring the
instrument's response to a known input), or by a combination of theory and
experiment (by measuring parameters of an instrument model). In all cases the
accuracy of the restored fluorophore distribution is limited by the accuracy of
the imaging model. All theoretical imaging models make simplifying
approximations that limit the accuracy of restorations, so it is important to
verify that the approximations introduce an acceptable level of error. This work
investigates the errors introduced by two common approximations in models of
fluorescence microscopes---the \textit{monopole approximation} and the
\textit{scalar approximation}.

The widely-used monopole model treats fluorophores as molecules that absorb and
emit light isotropically. Despite their use in models, monopole
absorbers/emitters do not exist in nature. All real fluorophores absorb and emit
light anisotropically, and in this work we investigate the more realistic dipole
model of fluorophores. Since the absorption and emission patterns of many widely
used fluorophores including green fluroescent protein (GFP) are known to be well
described by the dipole model [cite Inoue], we start by considering the dipole
model before higher-order multipole models.

The scalar approximation is another widely-used approximation in fluorescence
microscopy. The scalar approximation models the propagation of light using a
scalar-valued field $U$ instead of the more realistic vector-valued electric
field $\mb{E}$. Although the scalar model yields realistic results in many
cases, the scalar field ignores polarization-dependent effects. The absorption
and emission response of dipoles strongly depends on the polarization of the
fields, so we need to use an electromagnetic model of light propagation to
accurately model a microscope imaging dipoles.

This work lies at the intersection of three classes of fluorescence microscopy:
(1) ensemble imaging techniques where many fluorophores are imaged in parallel
and the goal is to find their concentration throughout the sample, (2) single
molecule localization microscopy (SMLM) techniques where fluorophores are sparse
in the sample and the goal is to find the position of each fluorophore, and (3)
polarized fluorescence techniques where many fluorophores are imaged in parallel
and the goal is to find their concentration and orientation. We briefly review
these three classes and focus on their use of the monopole and scalar
approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and many widely used restoration techniques make either or both of
the monopole and scalar approximations. The Gibson-Lanni model \cite{gibson89}
and the Born and Wolf model \cite{born1980} are two examples of scalar models
that are widely used to restore data collected by fluorescence microscopes, and
the Richards and Wolf model \cite{richards} is another widely-used model that
accounts for vectorial diffraction. All three of these models describe how
optical fields propagate through an optical system, but they are agnostic to the
object that is being imaged. In other words, these models are \textit{optical}
models, and when they are used to model fluorescent objects they implicitly make
the monopole approximation.

The SMLM community has pioneered the use of rigorous electromagnetic models for
dipole imaging systems that do not adopt the scalar or monopole approximations
\cite{backer2014, lieb2004, nov2006}. When a single molecule fluoresces, its
irradiance pattern on the detector is strongly dependent on the orientation of
the emitting molecule. Backlund et al. \cite{backlund2014} have shown have shown
that ignoring the orientation of fluorophores can bias position estimates, so
the most accurate SMLM experiments must jointly estimate the position and
orientation of each molecule. Some SMLM experiments use polarizers to estimate
the orientation of single molecules more accurately [Backer DNA paper, other
Moerner papers], and these works do not adopt the scalar or monopole
approximations.

A smaller community of microscopists is interested in measuring the orientation
and position of ensembles of fluorophores \cite{mehta2016} [Forkey, Goldman,
Moerner, Oldenbourg]. These techniques typically use polarized illumination or
polarized detection to make multiple measurements of the same object. Current
angular restoration techniques use a model of the dipole excitation and emission
processes [Fourkas] to recover the orientation of fluorophores using pixel-wise
arithmetic, but these techniques do not perform any spatial restoration so they
do not use all of the available information.

In this work we place these three communities on a common theoretical footing.
We begin in Section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In Section \ref{sec:results} we use the mathematical
tools we developed in Section \ref{sec:theory} to analyze and simulate a
single-view epi-fluorescence microscope, and we demonstrate our model with four
simple phantoms. Finally, in Section \ref{sec:discussion} we discuss the results
and their broader implications.

\section{Theory}\label{sec:theory}
We begin our analysis with the abstract Hilbert space formalism of Barrett and
Myers \cite{barrett2004}. Our first task is to formulate the imaging process as
a mapping between two Hilbert spaces $\mc{H}: \mbb{U} \rightarrow \mbb{V}$, where
$\mbb{U}$ is a set than contains all possible objects, $\mbb{V}$ is a set that
contains all possible datasets, and $\mc{H}$ is a model of the instrument that
maps between these two spaces. We denote (possibly infinite-dimensional)
Hilbert-space vectors in $\mbb{U}$ with $\mb{f}$, Hilbert-space vectors in
$\mbb{V}$ with $\mb{g}$, and the mapping between the spaces with
\begin{align}
  \mb{g} = \mc{H}\mb{f}.
\end{align}

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to unify the analysis of spatio-angular fluorescence microscopes.
In Section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In Section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

Next, we name the representations of our object and data in a specific basis. In
a delta function basis the object can be represented by a function $f(\ro)$
called the \textit{monopole density}---the number of monopoles per unit area at
the two-dimensional position $\ro$. Similarly, in a delta function basis the
data can be represented by a function $g'(\rd')$ called the
\textit{irradiance}---the power received by a surface per unit area at position
$\rd'$. It may seem pedantic to emphasize that these representations are in a
delta function basis, but we highlight this point to make it clear that the
delta function basis is not special---we can choose a different basis at will.

A reasonable starting point is to assume that the relationship between the
object and the data is \textit{linear}---this is usually true in fluorescence
microscopes since fluorophores emit incoherently, so a scaled sum of
fluorophores will result in a scaled sum of the irradiance patterns created by
the individual fluorophores. If the mapping is linear, we can write the
irradiance as a weighted integral over a field of monopoles
\begin{align}
g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd',\ro)f(\ro), \label{eq:fwdmono}
\end{align}
where $h'(\rd{}', \ro{})$ is the irradiance at position $\rd'$ created by a
point source at $\ro$.

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite{mansuripur2009}, so the imaging system can be modeled
within the field-of-view of the optical system as a magnifier with
shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor.

We can simplify our analysis by changing coordinates and writing Eq.
\ref{eq:nonconv} as a convolution \cite{barrett2004}. We define a demagnified
detector coordinate $\rd = \rd'/m$ and a normalization factor that corresponds
to the total power incident on the detector plane due to a point source
$P_{\text{mono}} = \int_{\mbb{R}^2}d\mb{r}\,h'(m\mb{r})$ where
$\mb{r} = \rd - \ro$. We use these scaling factors to define the
\textit{monopole point spread function} as
\begin{align}
  h(\rd - \ro) = \frac{h'(m[\rd - \ro])}{P_{\text{mono}}},
\end{align}
and the \textit{scaled irradiance} as
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_{\text{mono}}}.
\end{align}
With these definitions we can express the mapping between the object and the
data as a familiar convolution
\begin{align}
  g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro).  \label{eq:lsi}
\end{align}
We will mostly be using scaled coordinates, so we have adopted a slightly
unusual convention of using primes to denote unscaled coordinates.

@@

The mapping between the object and the data in a linear shift-invariant imaging
system takes a particularly simple form in a spatial harmonic basis. If we apply
the Fourier convolution theorem to Eq. \ref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real, so we know
that the monopole transfer function is normalized $H(0) = 1$ and
conjugate symmetric $H(-\bv) = H^*(\bv)$ where $z^*$ denotes the complex
conjugate of $z$.

Notice that Eqs. \ref{eq:lsi} and \ref{eq:freq} are expressions of the same
mapping between object and data space in different bases. Figure
\ref{fig:monopole-block} summarizes the relationship between object and data
space in both bases.

@@

Since the monopole transfer function is given by the two-dimensional Fourier
transform of the monopole point spread function
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\,h(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\bv],
\end{align}
we can plug in Eq. \ref{eq:absquarescalar} and use the autocorrelation theorem to
rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is often straightforward to calculate the field in a
Fourier plane of the detector. A typical approach for calculating the transfer
functions is to (1) calculate the field in a Fourier plane of the detector, (2)
scale the field to find the monopole coherent transfer function, then (3) use
the relationships in Figure \ref{fig:monopole-transfer-functions} to calculate
the other transfer functions.

@@

\subsection{Dipole imaging in different bases}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus dipoles by recording
the irradiance on a two-dimensional detector. A function that assigns a real
number to each point on a plane is not sufficient to specify a field of dipoles
because the dipoles can have different orientations. To represent the object we
need to extend object space to $\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$
where $\mbb{S}^2$ is the two-dimensional sphere (the usual sphere embedded in
$\mbb{R}^3$).

>>>>

% Talon's custom commands
\setlength\arraycolsep{2.25pt}
\let\originalleft\left
\let\originalright\right
\renewcommand{\left}{\mathopen{}\mathclose\bgroup\originalleft}
\renewcommand{\right}{\aftergroup\egroup\originalright}
\DeclareMathAlphabet{\mathcal}{OMS}{cmsy}{m}{n}
\newcommand{\tensor}[1]{\overset{\text{\tiny$\leftrightarrow$}}{\mb{#1}}}
\newcommand{\argmin}{\arg\!\min}
\newcommand{\me}{\mathrm{e}}
\providecommand{\e}[1]{\ensuremath{\times 10^{#1}}} 
\providecommand{\mb}[1]{\mathbf{#1}}
\providecommand{\msf}[1]{\mathsf{#1}}
\providecommand{\mf}[1]{\mathfrak{#1}}
\providecommand{\mc}[1]{\mathcal{#1}}
\providecommand{\ro}{\mathbf{\mathbf{r}}_o}
\providecommand{\so}{\mathbf{\hat{s}}_o}
\providecommand{\rp}{\mathbf{r}_p}
\providecommand{\rbm}[1]{r_b^{\text{m}}}
\providecommand{\rd}{\mathbf{r}_d}
\providecommand{\rdf}{\mathpzc{r}_d}
\providecommand{\mh}[1]{\mathbf{\hat{#1}}}
\providecommand{\mbb}[1]{\mathbb{#1}}
\providecommand{\bs}[1]{\boldsymbol{#1}}
\providecommand{\bv}{\bs{\nu}}
\providecommand{\taup}{\bs{\tau}}
\providecommand{\bsh}[1]{\hat{\boldsymbol{#1}}}
\providecommand{\nan}{\left(\frac{\text{NA}}{n_o}\right)}
\providecommand{\lmsum}{\sum_{\ell=0}^\infty\sum_{m=-\ell}^{\ell}}
\providecommand{\intr}[1]{\int_{\mbb{R}^{#1}}}
\providecommand{\ints}[1]{\int_{\mbb{S}^{#1}}}
\providecommand{\tb}[1]{\textcolor{blue}{#1}}
% \DeclareFontFamily{OT1}{pzc}{}
% \DeclareFontShape{OT1}{pzc}{m}{it}{<-> s * [1.10] pzcmi7t}{}
% \DeclareMathAlphabet{\mathpzc}{OT1}{pzc}{m}{it}
\newcommand{\eqname}[1]{\tag*{#1}}
\newcommand*\widefbox[1]{\fbox{\hspace{1em}#1\hspace{1em}}}

@@

\author{Talon Chandler,\authormark{1,*}, Rudolf Oldenbourg,\authormark{2} and\\ Patrick La Rivi\`ere\authormark{1,2}}

\address{\authormark{1}University of Chicago, Department of Radiology, Chicago, Illinois 60637, USA\\
  \authormark{2}Marine Biological Laboratory, Bell Center, Woods Hole, Massachusetts 02543, USA
}

@@

\begin{abstract*}
  We introduce the basic elements of a spatio-angular theory of fluorescence
  microscopy. We start by modeling an aplanatic microscope imaging an ensemble
  of in-focus fluorescent dipoles as a linear Hilbert-space operator, and we
  show that the operator takes a particularly convenient form when expressed in
  a basis of spatial and spherical harmonics---a form we call the dipole
  spatio-angular transfer function. We demonstrate our formalism by analyzing a
  single-view fluorescence microscope without using the monopole or scalar
  approximations. We show that this imaging system has a spatio-angular band
  limit, and we exploit the band limit to perform efficient simulations.
  Notably, we show that information about the out-of-plane orientation of
  ensembles of in-focus fluorophores is recorded by paraxial fluorescence
  microscopes.
\end{abstract*}

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring the spatial and angular distribution of ensembles of fluorophores.
While most fluorescence microscopy experiments are designed to measure only the
spatial distribution of fluorophores, a growing number of experiments use
polarizers to measure both the spatial and angular distributions of fluorophores
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}.

Meanwhile, single-molecule localization microscopy (SMLM) experiments use
spatially sparse fluorescent samples to localize single molecules with precision
that surpasses the diffraction limit. Real SMLM experiments are limited by noise
\cite{foreman2011, chao2016} and model mismatch, and several studies have shown
that model mismatch (e.g. ignoring the effects of vector optics, dipole
orientation \cite{backlund2014}, and dipole rotation \cite{lew2013}) can
introduce localization bias into SMLM experiments. Therefore, the most precise
SMLM experiments must use an appropriate model and jointly estimate both the
position and orientation of each fluorophore. Several studies have successfully
used vector optics and dipole models to estimate the position and orientation of
single molecules \cite{bohmer2003, lieb2004, toprak2006, aguet2009}, and there
is growing interest in designing optical systems for measuring the position,
orientation, and rotational dynamics of single molecules \cite{agrawal2012,
  backer2014, zhang2018, zhang2018-2}.

While many studies have focused on improving imaging models for spatially sparse
fluorescent samples, we consider the more general case and aim to improve
imaging models for arbitrary samples. In particular, we examine the effects of
two widely used approximations in fluorescence microscopy---the \textit{monopole
  approximation} and the \textit{scalar approximation}.

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think that it is an accurate label
for the way many models of fluorescence microscopes treat fluorophores, and we
use the term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, monopole absorber/emitters do
not exist in nature, and the monopole approximation must be justified. Although
many works implicitly apply the monopole approximation, we have encountered two
explicit justifications: (1) the sample contains many randomly oriented
fluorophores within a resolvable volume or (2) the sample contains unconstrained
rotating fluorophores. While both of these situations yield monopole-like
samples, neither yields a sample that is perfectly described by the monopole
model. In this work we investigate the dipole model of fluorophores in detail
and find the conditions under which the monopole approximation is justified.

All fluorescence microscopy techniques that use an \textit{optical point spread
  function} or an \textit{optical transfer function} to model the mapping
between the fluorophores and the irradiance on the detector implicitly make the
monopole approximation. The optical point spread function is the irradiance
response of an optical system to an isotropic point source, so it cannot model
the response due to an anisotropic dipole radiator. In this work we define
\textit{monopole and dipole transfer functions} that describe the mapping
between fluorophores and the measured irradiance. Although optical systems are
an essential part of microscopes, fluorescence microscopists are interested in
measuring fluorophores (not optics), so the monopole and dipole transfer
functions are more valuable than the optical transfer function.

While the monopole approximation applies to the fluorescent object, the
\textit{scalar approximation} applies to the fields that propagate through the
microscope. Modeling the electric fields in a region requires a
three-dimensional vector field, but if the electric fields are random or
parallel a scalar field is sufficient. Replacing the vector-valued electric
field, $\mb{E}$, with a scalar-valued field, $U$, is the scalar approximation,
and this approximation is only justified if the fields are randomly or uniformly
polarized.

The scalar approximation is often made together with the monopole approximation.
For example, the Born-Wolf model \cite{born1980} and the Gibson-Lanni model
\cite{gibson89} make both the monopole and scalar approximations when applied to
fluorescence microscopes. However, some models make the monopole approximation
but not the scalar approximation. For example, the Richards-Wolf model
\cite{richards} considers the role of vector-valued fields in the optical
system, but it is an optical model so when it is applied to fluorescence
microscopes the monopole approximation is required. As previously mentioned,
several works in the SMLM community have modeled and applied dipole point spread
functions that do not make the monopole or scalar approximations, but to our
knowledge no works have considered dipole transfer functions.

This work lies at the intersection of three fluorescence microscopy communities:
(1) the spatial ensemble imaging community where many fluorophores are imaged in
parallel and the goal is to find their concentration throughout the sample, (2)
the spatio-angular ensemble community where many fluorophores are imaged in
parallel and the goal is to find their concentration and orientation, and (3)
the single molecule localization microscopy (SMLM) community where fluorophores
are sparse in the sample and the goal is to find the position of each
fluorophore. We briefly review how these three communities use the monopole and
scalar approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and all existing restoration techniques make either or both of the
monopole and scalar approximations. As discussed above, the Gibson-Lanni model,
the Born-Wolf model, and the Richards-Wolf model are approximate when applied to
fluorescence microscopes because they only model monopole emitters.
Deconvolution algorithms that use these models may make biased estimates of
fluorophore concentrations since they ignore the dipole excitation and emission
of fluorophores, but we will see that these effects are small when the sample
contains many randomly oriented fluorophores within a resolvable volume or
unconstrained rotating fluorophores.

A small but growing group of microscopists is interested in measuring the
orientation and position of ensembles of fluorophores \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017, zhanghao2017}. These techniques
typically use polarizers to make multiple measurements of the same object, then
they use a model of the dipole excitation and emission processes
\cite{fourkas2001} to recover the orientation of fluorophores using pixel-wise
arithmetic. Although these studies do not adopt the scalar or monopole
approximations for the angular part of the problem, they adopt both
approximations when they consider the spatial part of the problem. Existing
works either ignore the spatial reconstruction problem \cite{vrabioiu2006,
  mattheyses2010, mehta2016, mcquilken2017} or assume that the spatial and
angular reconstruction problems can be performed sequentially
\cite{zhanghao2017}.

The most precise experiments in the SMLM community do not adopt the scalar or
monopole approximations. Although many works have applied dipole models with
vector optics, fewer have considered the effects of rotational or spatial
diffusion, and to our knowledge no works have considered both rotational and
spatial diffusion together. We will see that the dipole transfer functions are
useful tools for incorporating angular and spatial diffusion into SMLM
simulations and reconstructions.

In this work we place these three communities on a common theoretical footing.
We begin in section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In section \ref{sec:results} we use the mathematical
tools we developed in section \ref{sec:theory} to analyze and simulate a
single-view fluorescence microscope, and we demonstrate our techniques with four
simple phantoms. Finally, in section \ref{sec:discussion} we discuss the results
and their broader implications.

\section{Theory}\label{sec:theory}
We begin our analysis with the abstract Hilbert space formalism of Barrett and
Myers \cite[ch.~1.3]{barrett2004}. Our first task is to formulate
the imaging process as a mapping between two Hilbert spaces
$\mc{H}: \mbb{U} \rightarrow \mbb{V}$, where $\mbb{U}$ is a set that contains
all possible objects, $\mbb{V}$ is a set that contains all (possibly
noise-corrupted) datasets, and $\mc{H}$ is a model of the instrument that maps
between these two spaces. We denote (possibly infinite-dimensional)
Hilbert-space vectors in $\mbb{U}$ with $\mb{f}$, Hilbert-space vectors in
$\mbb{V}$ with $\mb{g}$, and the mapping between the spaces with
\begin{align}
  \mb{g} = \mc{H}\mb{f}.
\end{align}
Throughout this work we will use the letters $g$, $h$, and $f$ with varying
fonts, capitalizations, and arguments to represent the data, the instrument, and
the object, respectively.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to unify the analysis of spatio-angular fluorescence microscopes.
In section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

Next, we name the representations of our object and data in a specific basis. In
a delta function basis the object can be represented by a function $f(\ro)$
called the \textit{monopole density}---the number of monopoles per unit area at
the two-dimensional position $\ro$. Similarly, in a delta function basis the
data can be represented by a function $g'(\rd')$ called the
\textit{irradiance}---the power received by a surface per unit area at position
$\rd'$. Note that we have adopted a slightly unusual convention of using primes
to denote unscaled coordinates. Later in this section we will introduce unprimed
scaled coordinates that we will use throughout the rest of the
paper.

A reasonable starting point is to assume that the relationship between the
object and the data is \textit{linear}---this is true in many fluorescence
microscopes because fluorophores emit incoherently, so a scaled sum of
fluorophores will result in a scaled sum of the irradiance patterns created by
the individual fluorophores. Note that our assumption of linearity excludes
cases where fluorophores interact (e.g. homoFRET) or saturate (e.g. non-linear
fluorescence microscopy).

If the mapping is linear, we can write the irradiance as a weighted integral
over a field of monopoles
\begin{align}
g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd',\ro)f(\ro), \label{eq:fwdmono}
\end{align}
where $h'(\rd', \ro{})$ is the irradiance at position $\rd'$ created by a
point source at $\ro$.

Next, we assume that the optical system is \textit{aplanatic}---Abbe's sine
condition is satisfied and on-axis points are imaged without aberration. Abbe's
sine condition guarantees that off-axis points are imaged without spherical
aberration or coma \cite[ch.~1]{mansuripur2009}, so the imaging system can
be modeled within the field of view of the optical system as a magnifier with
shift-invariant blur
\begin{align}
  g'(\rd') = \int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro)f(\ro), \label{eq:nonconv}
\end{align}
where $m$ is a magnification factor.

We can simplify our analysis by changing coordinates and writing Eq.
\eqref{eq:nonconv} as a convolution \cite[ch.~7.2.7]{barrett2004}. We
define a demagnified detector coordinate $\rd = \rd'/m$ and a normalization
factor that corresponds to the total power incident on the detector plane due to
a point source $P_{\text{mono}} = \int_{\mbb{R}^2}d\mb{r}\,h'(m\mb{r})$ where
$\mb{r} = \rd - \ro$. We use these scaling factors to define the
\textit{monopole point spread function} as
\begin{align}
  h(\rd - \ro) = \frac{h'(m[\rd - \ro])}{P_{\text{mono}}},
\end{align}
and the \textit{scaled irradiance} as
\begin{align}
  g(\rd) = \frac{g'(m\rd)}{P_{\text{mono}}}.
\end{align}
With these definitions we can express the mapping between the object and the
data as a familiar convolution
\begin{align}
  g(\rd) = \int_{\mbb{R}^2}d\ro\, h(\rd - \ro)f(\ro).  \label{eq:lsi}
\end{align}

@@

The mapping between the object and the data in a linear shift-invariant imaging
system takes a particularly simple form in a spatial harmonic (i.e. Fourier) basis. If we apply
the Fourier convolution theorem to Eq. \eqref{eq:lsi} we find that
\begin{align}
  G(\bv) = H(\bv)F(\bv),\label{eq:freq}
\end{align}
where we define the \textit{scaled irradiance spectrum} as
\begin{align}
  G(\bv) = \int_{\mbb{R}^2}d\mb{r}\, g(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),
\end{align}
the \textit{monopole transfer function} as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\mb{r}\, h(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv),\label{eq:otf}
\end{align}
and the \textit{monopole spectrum} as
\begin{align}
    F(\bv) = \int_{\mbb{R}^2}d\mb{r}\, f(\mb{r})\, \text{exp}(-2\pi i\mb{r}\cdot\bv).
\end{align}
The monopole point spread function is normalized and real, so we know that the
monopole transfer function is normalized, $H(0) = 1$, and conjugate symmetric,
$H(-\bv) = H^*(\bv)$, where $z^*$ denotes the complex conjugate of $z$.

Notice that Eqs. \eqref{eq:lsi} and \eqref{eq:freq} are expressions of the same
mapping between object and data space in different bases. Figure
\ref{fig:monopole-block} summarizes the relationship between object and data
space in both bases.

@@

We can plug Eq. \eqref{eq:absquarescalar} into Eq. \eqref{eq:otf} and use the
autocorrelation theorem to rewrite the monopole transfer function as
\begin{align}
  H(\bv) = \int_{\mbb{R}^2}d\taup\, C(\taup)C^*(\taup - \bv), 
\end{align}
where we have introduced the
\textit{monopole coherent transfer function} as the two-dimensional Fourier
transform of the monopole coherent spread function
\begin{align}
  C(\taup) = \int_{\mbb{R}^2}d\mb{r}\, c(\mb{r})\,\text{exp}[-2\pi i\mb{r}\cdot\taup].
\end{align}
Physically, the monopole coherent transfer function corresponds to the
scalar-valued field in a Fourier plane of the detector with appropriate scaling.
% We have chosen the symbol $\taup$ as the coordinate for the coherent
% transfer function because the Fourier plane of the detector is often called the
% \textit{pupil plane}.

The coherent transfer function provides a valuable shortcut for analyzing
microscopes since it is often straightforward to calculate the field in a
Fourier plane of the detector. A typical approach for calculating the transfer
functions is to (1) calculate the field in a Fourier plane of the detector, (2)
scale the field to find the monopole coherent transfer function, then (3) use
the relationships in Fig. \ref{fig:monopole-transfer-functions} to calculate
the other transfer functions.

@@

\subsection{Dipole imaging in different bases}\label{sec:dipole}
Now we consider a microscope imaging a field of in-focus dipoles by recording
the irradiance on a two-dimensional detector. A function that assigns a real
number to each point on a plane is not sufficient to specify a field of dipoles
because the dipoles can have different orientations. To represent the object we
need to extend object space to
$\mbb{U} = \mbb{L}_2(\mbb{R}^2\times\mbb{S}^2)$---the set of square-integrable
functions on the product space of a plane and a two-dimensional sphere (the
usual sphere embedded in $\mbb{R}^3$). To visualize functions in object space we
imagine a sphere at every point on a plane with a scalar value assigned to every
point on each sphere.

In a delta function basis the object can be represented by a function
$f(\ro, \so)$ called the \textit{dipole density}---the number of dipoles per
unit area and per unit solid angle at position $\ro{}$ and oriented along
$\so{}$. Similar to the monopole case, we model the mapping between the object
and the irradiance in a delta function basis as an integral transform
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd', \ro, \so)f(\ro, \so),
\end{align}
where $h'(\rd', \ro, \so)$ is the irradiance at position $\rd'$ created by a
point source at $\ro$ with orientation $\so$. If the optical system is
aplanatic, we can write the integral transform as
\begin{align}
  g'(\rd') = \int_{\mbb{S}^2}d\so\int_{\mbb{R}^2}d\ro\, h'(\rd' - m\ro, \so)f(\ro, \so). 
\end{align}
